Staged Abstract Interpreters

### 2019-09-04 paper/20_prelim.tex
revert to as sign; typo
revision

In this section, we first describe the abstract syntax of the language for our interpreters. 
Then present the generic interpreter shared among the four different
semantics, after which, we instantiate the interpreter to the concrete one.
It is worth noting that we choose to use Scala and monadic style to
demonstrate the idea, but the approach is not restricted to our
choice. One can use imperative or direct style in other MSP languages
(e.g., MetaOCaml \cite{DBLP:conf/gpce/CalcagnoTHL03, DBLP:conf/flops/Kiselyov14}
and Template Haskell \cite{Sheard:2002:TMH:636517.636528} to construct
such staged abstract interpreters.

@@

We consider a call-by-value $\lambda$-calculus in direct-style, extended
with numbers, arithmetic, recursions, and conditionals. Other effectful features
such as assignments can also be supported readily.
%In Section~\ref{cases_imp}, we will add more imperative features to the language.
Since we are mostly interested in analyzing the dynamic behaviors of the
program, we disguise any static semantics and type system. We also assume that
input programs are well-typed and all variables are distinct. The abstract
syntax is shown as follows:

\begin{lstlisting}
  abstract class Expr
  case class Lit(i: Int) extends Expr                         // numbers
  case class Var(x: String) extends Expr                      // variables
  case class Lam(x: String, e: Expr) extends Expr             // abstractions
  case class App(e1: Expr, e2: Expr) extends Expr             // applications
  case class If0(e1: Expr, e2: Expr, e3: Expr) extends Expr   // conditionals
  case class Rec(x: String, rhs: Expr, e: Expr) extends Expr  // recursions
  case class Aop(op: String, e1: Expr, e2: Expr) extends Expr // arithmetic
\end{lstlisting}

@@

\begin{figure}[h!]
  \centering
  \begin{subfigure}[b]{0.55\textwidth}
    \begin{lstlisting}
  trait Monad[M[_]] {                                  
    def pure[A](a: A): M[A]                            
    def flatMap[A,B](ma: M[A])(f: A => M[B]): M[B]     
  }                                                    
    \end{lstlisting}
    \caption{trait \texttt{Monad}} \label{fig:monad}
  \end{subfigure}
  ~
  \begin{subfigure}[b]{0.4\textwidth}
    \begin{lstlisting}
trait MonadOps[M[_], A] {
  def map[B](f: A => B): M[B]
  def flatMap[B](f: A => M[B]): M[B]
}
    \end{lstlisting}
    \caption{trait \texttt{MonadOps}} \label{fig:monadops}
  \end{subfigure}
\end{figure}

Similar to Haskell's @do@-notation, Scala provides special syntactic support for
monadic operations through @for@-comprehension.
For example, an object of @List[A]@ is an instance of @List@ monad, where @A@ is the element type. 
Then to compute the Cartesian product of two lists of numbers, we can use Scala's
@for@-comprehension syntax.
\begin{lstlisting}
  val xs = List(1, 2); val ys = List(4, 5)
  for { x <- xs; y <- ys } yield (x, y) // List((1,4), (1,5), (2,4), (2,5))
\end{lstlisting}

@@

Other primitive operations provide basic functionality for the
interpreter.  The method @num@ and @close@ deal with primitive values,
which lift literal terms (e.g., lambdas) to our value representation
(e.g., closures).  The method @get@ retrieves the value of an
identifier @x@, through the environment and store. Conditionals and
arithmetic is handled by @br0@ and @arith@, respectively. The method
@ap_clo@ is used for function applications: it takes a function value
and an argument value. Note that the @Env@, @Store@, and @Value@ are
all annotated by @R[_]@ since they can potentially be next-stage
values when acting as compilers.
\begin{lstlisting}
  def num(i: Int): Ans
  def get(σ: R[Store], ρ: R[Env], x: Ident): R[Value]
  def close(ev: Expr => Ans)(λ: Lam, ρ: R[Env]): R[Value]
  def br0(test: R[Value], thn: => Ans, els: => Ans): Ans
  def arith(op: Symbol, v1: R[Value], v2: R[Value]): R[Value]
  def ap_clo(ev: Expr => Ans)(rator: R[Value], rand: R[Value]): Ans
\end{lstlisting}

\begin{figure}[h!]
  \centering
  \begin{lstlisting}
          def eval(ev: Expr => Ans)(e: Expr): Ans = e match {
            case Lit(i) => num(i)                   case Let(x, rhs, e) => for {
            case Var(x) => for {                      v  <- ev(rhs)
              ρ <- ask_env                            ρ  <- ask_env
              σ <- get_store                          α  <- alloc(x)
            } yield get(σ, ρ, x)                      _  <- set_store(α → v)
            case Lam(x, e) => for {                   rt <- local_env(ev(e))(ρ + (x → α))
              ρ <- ask_env                          } yield rt
            } yield close(ev)(Lam(x, e), ρ)         case Aop(op, e1, e2) => for {
            case App(e1, e2) => for {                 v1 <- ev(e1)                                               
              v1 <- ev(e1)                            v2 <- ev(e2)
              v2 <- ev(e2)                          } yield arith(op, v1, v2)
              rt <- ap_clo(ev)(v1, v2)              case Rec(x, rhs, e) => for {
            } yield rt                                α  <- alloc(x)
            case If0(e1, e2, e3) => for {             ρ  <- ask_env
              cnd <- ev(e1)                           v  <- local_env(ev(rhs))(ρ + (x → α))
              rt  <- br0(cnd, ev(e2), ev(e3))         _  <- set_store(α → v)
            } yield rt                                rt <- local_env(ev(e))(ρ + (x → α))
                                                    } yield rt                    
          }
  \end{lstlisting}
\caption{The generic interpreter,
  shared by the unstaged/staged + concrete/abstract interpreter.}
\label{fig:shared_int}
\end{figure}

>>>>

In this section, we first describe the abstract syntax of the language for our interpreters. 
Then we present the generic interpreter shared among the four different
semantics. After which, we instantiate the interpreter to the concrete one.
It is worth noting that we choose to use Scala and monadic style to
demonstrate the idea, but the approach is not restricted to this
choice. One can use imperative or direct style in other MSP languages
(e.g., MetaOCaml \cite{DBLP:conf/gpce/CalcagnoTHL03, DBLP:conf/flops/Kiselyov14}
and Template Haskell \cite{Sheard:2002:TMH:636517.636528} to construct
such staged abstract interpreters.

@@

We consider a call-by-value $\lambda$-calculus in direct-style, extended
with numbers, arithmetic, recursions, and conditionals. Other effectful features
such as assignments can also be supported readily.
Since we are mostly interested in analyzing the dynamic behaviors of the
program, we will disguise the static semantics. We also assume that
input programs are well-typed and all variables are distinct. The abstract
syntax is shown as follows:
\begin{lstlisting}
  abstract class Expr
  case class Lit(i: Int) extends Expr                         // numbers
  case class Var(x: String) extends Expr                      // variables
  case class Lam(x: String, e: Expr) extends Expr             // abstractions
  case class App(e1: Expr, e2: Expr) extends Expr             // applications
  case class If0(e1: Expr, e2: Expr, e3: Expr) extends Expr   // conditionals
  case class Rec(x: String, rhs: Expr, e: Expr) extends Expr  // recursions
  case class Aop(op: String, e1: Expr, e2: Expr) extends Expr // arithmetic
\end{lstlisting}

@@

\begin{figure}[h!]
  \centering
  \begin{subfigure}[b]{0.55\textwidth}
    \begin{lstlisting}
  trait Monad[M[_]] {                                  
    def pure[A](a: A): M[A]                            
    def flatMap[A,B](ma: M[A])(f: A => M[B]): M[B]     
  }                                                    
    \end{lstlisting}
    \caption{trait \texttt{Monad}} \label{fig:monad}
  \end{subfigure}
  ~
  \begin{subfigure}[b]{0.4\textwidth}
    \begin{lstlisting}
trait MonadOps[M[_], A] {
  def map[B](f: A => B): M[B]
  def flatMap[B](f: A => M[B]): M[B]
}
    \end{lstlisting}
    \caption{trait \texttt{MonadOps}} \label{fig:monadops}
  \end{subfigure}
\end{figure}
\vspace{-1em}

Similar to Haskell's @do@-notation, Scala provides special syntactic support for
monadic operations through @for@-comprehension.
For example, an object of @List[A]@ is an instance of @List@ monad, where @A@ is the element type. 
Then to compute the Cartesian product of two lists of numbers, we can use Scala's
@for@-comprehension syntax.
\begin{lstlisting}
  val xs = List(1, 2); val ys = List(3, 4)
  for { x <- xs; y <- ys } yield (x, y)   // List((1,3), (1,4), (2,3), (2,4))
\end{lstlisting}

@@

Other primitive operations provide basic functionality for the
interpreter.  The method @num@ and @close@ deal with primitive values,
which lift literal terms (e.g., lambdas) to our value representation
(e.g., closures).  The method @get@ retrieves the value of an
identifier @x@, through the environment and store. Conditionals and
arithmetic are handled by @br0@ and @arith@, respectively. The method
@ap_clo@ is used for function applications: it takes a function value
and an argument value. Note that the @Env@, @Store@, and @Value@ are
all annotated by @R[_]@ since they can potentially be next-stage
values when acting as compilers.
\begin{lstlisting}
  def num(i: Int): Ans
  def get(σ: R[Store], ρ: R[Env], x: Ident): R[Value]
  def close(ev: Expr => Ans)(λ: Lam, ρ: R[Env]): R[Value]
  def br0(test: R[Value], thn: => Ans, els: => Ans): Ans
  def arith(op: Symbol, v1: R[Value], v2: R[Value]): R[Value]
  def ap_clo(ev: Expr => Ans)(rator: R[Value], rand: R[Value]): Ans
\end{lstlisting}

\begin{figure}[h!]
  \centering
  \begin{lstlisting}
          def eval(ev: Expr => Ans)(e: Expr): Ans = e match {
            case Lit(i) => num(i)                   case Let(x, rhs, e) => for {
            case Var(x) => for {                      v  <- ev(rhs)
              ρ <- ask_env                            ρ  <- ask_env
              σ <- get_store                          α  <- alloc(x)
            } yield get(σ, ρ, x)                      _  <- set_store(α → v)
            case Lam(x, e) => for {                   rt <- local_env(ev(e))(ρ + (x → α))
              ρ <- ask_env                          } yield rt
            } yield close(ev)(Lam(x, e), ρ)         case Aop(op, e1, e2) => for {
            case App(e1, e2) => for {                 v1 <- ev(e1)                                               
              v1 <- ev(e1)                            v2 <- ev(e2)
              v2 <- ev(e2)                          } yield arith(op, v1, v2)
              rt <- ap_clo(ev)(v1, v2)              case Rec(x, rhs, e) => for {
            } yield rt                                α  <- alloc(x)
            case If0(e1, e2, e3) => for {             ρ  <- ask_env
              cnd <- ev(e1)                           v  <- local_env(ev(rhs))(ρ + (x → α))
              rt  <- br0(cnd, ev(e2), ev(e3))         _  <- set_store(α → v)
            } yield rt                                rt <- local_env(ev(e))(ρ + (x → α))
                                                    } yield rt                    
          }
  \end{lstlisting}
\caption{The generic interpreter,
  shared by the unstaged/staged + concrete/abstract interpreter.}
\label{fig:shared_int}
\end{figure}
\vspace{-1em}
